\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-10-27 metatheory-condensed.tex
Comments on how linkage already capture code reuse

Now, we will mainly focus on the newly introduced typing rules, and illustrate their connection to the plugin implementation. For those curious about the complete meta-theory formulation, we refer to our appendices or \citet{altkap2016, kaposi2017type} or maybe those not using QIIT \cite{coquand2018canonicity, sterling2019algebraic}. Our metatheory is formulated using QIIT in an extensional setting.
%\YZ{
%    Is extensionality in conflict with the claim in §3 that "subst tm_unit x t" and "tm_unit" are
%    not definitionally equal but propositionally equal in family STLC?
%}\EDJreply{No, TLDR: it is a different level. \\ 
%There are two type theories, one is Agda-like (= Latex = Conventional math), the other is the one formulated in QIIT. This sentence is saying that the Agda-like one is extensional MLTT (i.e. we have function extensionality and etc here in Latex writing directly). \\ But the one you are asking about is in the metatheory in QIIT: if we have an identity type $Id x y$ (propositionally equal) can we have a judgemetnal equality for the two terms in QIIT $x \equiv y$. \\
%Apparently, we don't setup our QIIT formulation in this way (I thought we need, but now I don't think we need it. Our QIIT formulation is purely in intensional MLTT now).\\ 
%But even if we do setup like that, I will not say a conflict is happening (because it is to our advantage if we have judgemental equal in metatheory). I will say it is a bit off as to how Coq is formulated (Coq is closer to an intensional MLTT, so no judgemental equal only propositional one). 
%}

% linkage, signature and compilation and seal
$$
\Rule[name=lsig/ept]
{""}
{\goodSig{\Gm}{\LSigEmp}{0}}
\quad
\Rule[name=ty/lkg]
{\goodSig{\Gm}{\lsig}{n}}
{\goodType{\Gm}{\TyLkg{\lsig}}{i}
\\ \goodType{\Gm}{\TyTkg{\lsig}}{}}
\quad
\Rule[name=Lnkg Compile]
{ \goodTerm{\Gamma}{o}{\cL \sigma} 
}
{
  \goodTerm{\Gamma}{\Tkg {o}}{\cC \sigma}
}
$$
$$
\Rule[name=Sig Add]
{\goodSig{\Gamma}{\sigma}{n} 
 \quad \goodType{\Gamma}{A}{}
 \quad \goodSeal{\Gamma}{f}{\sigma}{A}
 \quad \goodType{\Gamma, A}{T}{i}}
{\goodSig{\Gamma}{(\nu^+ \ \sigma \ \{f\} \ T)}{n+1}}
\quad
\Biggl( \boxed{\goodSeal{\Gamma}{s}{\sigma}{A}} 
:= \boxed{\goodTerm{\Gamma, \cC\sigma}{s}{A[p^1]}} \Biggr)
$$
$$
\Rule[name=Ept Lnkg]
{}
{\goodTerm{\Gamma}{\mu\cdot}{\cL \nu\cdot}}
\quad
\Rule[name=Lnkg Add]
{ \goodTerm{\Gamma}{o}{\cL \sigma} 
\quad \goodType{\Gamma}{A}{}
\quad  {\goodSeal{\Gamma}{f}{\sigma}{A}}
 \quad \boxed{\goodTerm{\Gamma, A}{t}{T}}
}
{\goodTerm{\Gamma}{(\mu^+ \ o \ \{f\} \ t)}{\cL(\nu^+ \ \sigma \ \{f\} \ T)}}
$$
We start with signature (in a brand new judgement $\goodSig{\Gamma}{\sigma}{n}$). A signature is a list of types contained by a \textit{linkage}\footnote{Following the convention by \citet{zm2017}, we name the meta-theoretical correspondents of \textit{family} as \textit{linkage}.}, tracking the length as superscript.   We will use $\{a : T_1, b : T_2, ..\}$ as a shorthand for signature with type $T_1, T_2, ..$ for readability. 
\newpage 
Adding new types into signature \ruleref{Sig Add} requires a \textit{sealing judgement} $\goodSeal{\Gamma}{f}{\sigma}{A}$, which is an alias of a term judgement. Sealing is corresponding \textit{to indicating whether each field is overridable(extensible) or non-overridable (thus definitionally exposed) in the Coq Plugin}. 
Concretely, we encode our STLC example to show the motivation of the sealing.
%\YZ{Is sealing used only for late-binding the signatures of inductive types? Do I understand correctly that late binding of other kinds of fields (like subst) is achieved by the boxed premise in rule "LNKG ADD"?}\EDJreply{All late-bound are achieved by ``selective-abstraction'' imposed by sealing. For example, recursion will be consequently abstracted because of the abstraction of the inductive type it recurse on. 
%}\EDJreply{Sorry, I see what you mean. Yes. the box part is important for late-binding. But the functionality of sealing is always there. They are both required to achieve overriding. A illustrative example is, if the type is $S(\bot)$, then even with the boxed premise, it is still not overridable because the type itself restricts all the possibilities.}\YZreply{
%  By "selective abstraction", do you mean in the metatheory the programmer chooses what and how to abstract?
%}\EDJreply{Yes.}

In \cref{fig:sealing+linkage}, the "hiding" comment illustrates the late-bound prior fields for each fields. For "tm", "ty" is late-bound and thus extending "ty" won't cause disruption on the inheritance of "tm". For "subst" itself, the definition of "tm" is not hidden to achieve exhaustiveness checking. However, for each case handler inside "subst", "tm" is late-bound and thus each handler is inheritable.

Each hiding comment corresponds to the textual-aligned sealing derivation on RHS. We use surface syntax for eligibility: curly brackets indiate a signature (omitting sealing at type level for readability) and square bracket indicate a nested sigma type, corresponding to sealing judgement $\goodSeal{}{s}{\sigma}{A}$.

For "tm", "ty" is hidden and thus the sealing "s₂" will make "ty" into an arbitrary type "𝕌" and expose its constructor "bool", "arrow", and etc. Then when constructing "o₂", the newly added type "tm" as "⋆ : 𝕎τ₂" is constructed in the context where "ty : 𝕌", ignoring its original eliminator. (Because eliminator is only possible when "ty : 𝕎τ₁"). On the other hand, for "subst", only "ty" is hidden and thus the sealing "s₃" leave "tm : 𝕎τ₂" untouched and "Wrec" becomes possible when constructing "o₃".

Basically, the hidden prior fields---inductive type are sealed into arbitrary type with ``constructors only'', where the recursion is constructed without sealing the inductive type.  Generally speaking, each extensible inductive type are hidden for all the other fields except for the recursion on itself. Furthermore, normal fields (e.g. "t1, t2") can expose its definition to all in meta-theory by using singleton type. (e.g.  saying "t1 : S(tvar 0)" instead of the current "t1 : tm"); and then overridable fields are hidden for normal fields (by sealing "t1 : S(tvar 0)" into "t1 : tm"). Generally speaking, anything with explicit "hiding" decoration can be understood as sealing.

Due to the definition of sealing, we can see the importance of compilation type, as the result of a sealing is usually a (nested) sigma type instead of a linkage type. For example, $\TyLkg{\{a : S(\bot), b : a \to \bot\}}$ can only be sealed into $\TyTkg{\{a : \cU, b : a \to \bot\}}$ but never $\TyLkg{\{a : \cU, b : a \to \bot\}}$. Because the later cannot be inhabitted---by projecting the second field of the $ t : \TyLkg{\{a : \cU, b : a \to \bot\}}$, we can get a proof of bottom type.

@@

Our formulation of (extensible) inductive type is close to W-type (\ruleref{WSig-Add} requires a pair of types $A \vdash B$ as well), but with mulitple constructors. Thus we have a brand new judgement ${\goodWSig{\Gamma}{\tau}{n}}$ indicating signature of inductive type by its constructors. Unconventionally, a signature can induce a "universe" of types $\bW\tau$---in other words, $T : \bW\tau$ will be the inductive type instead of $\bW\tau$ itself. "Wsup" is used to construct a specific term of a given inductive type.

When programming via meta-theory, we usually seal the inductive type, only expose its constructors (as a mundane function member) and hide the eliminator by abstracting $T : \bW\tau$ into $"W"~T : \cU$. Then the later fields---those only rely on this interface with $\cU$---can be inherited to other contexts (e.g. using enriched inductive type), as long as the targeting contexts can be abstracted into this {interface} again.

@@

@@

% Inhertance judgement
In the plugin implementation, "Family"s are second class objects like Coq's "Module". What's more, a "Family" (with or without "extends" clause) can be considered as a standalone piece of code trying to \textit{inherit} an existent (but maybe empty) family. Thus we model them using \textit{inheritance judgement} $\goodInh{\Gamma}{h}{\sigma_1}{\sigma_2}$ instead of considering them as first class terms. The only way to use $h$ is by \ruleref{Inh} and transforming a linkage of signature $\sigma_1$ to another linkage of $\sigma_2$. Inheritance judgement can be inductively constructed via inheriting operations, overriding operations, extending operations and etc, corresponding to "Inherit", "Override" and "Extend" command  in our plugin implementation.

>>>>

Now, we will mainly focus on the newly introduced typing rules, and illustrate their connection to the plugin implementation. For those curious about the complete meta-theory formulation, we refer to our appendices or \citet{altkap2016, kaposi2017type} or maybe those not using QIIT \cite{coquand2018canonicity, sterling2019algebraic}. Our metatheory is formulated using QIIT in an extensional setting.

@@

We start with the formulation of (extensible) inductive type, which is close to W-type (\ruleref{WSig-Add} requires a pair of types $A \vdash B$ as well), but with mulitple constructors. Thus we have a brand new judgement ${\goodWSig{\Gamma}{\tau}{n}}$ indicating signature of inductive type by its constructors. Unconventionally, a signature can induce a "universe" of types $\bW\tau$---in other words, $T : \bW\tau$ will be the inductive type instead of $\bW\tau$ itself. "Wsup" is used to construct a specific term of a given inductive type.

% When programming via meta-theory, we usually seal the inductive type, only expose its constructors (as a mundane function member) and hide the eliminator by abstracting $T : \bW\tau$ into $"W"~T : \cU$. Then the following fields---those only rely on this interface with $\cU$---can be inherited to other contexts (e.g. using enriched inductive type), as long as the targeting contexts can be abstracted into this {interface} again.

@@

To introduce the concrete programming on inductive type in metatheory, we need to first introduce the formulation of families.

%\YZ{
%    Is extensionality in conflict with the claim in §3 that "subst tm_unit x t" and "tm_unit" are
%    not definitionally equal but propositionally equal in family STLC?
%}\EDJreply{No, TLDR: it is a different level. \\ 
%There are two type theories, one is Agda-like (= Latex = Conventional math), the other is the one formulated in QIIT. This sentence is saying that the Agda-like one is extensional MLTT (i.e. we have function extensionality and etc here in Latex writing directly). \\ But the one you are asking about is in the metatheory in QIIT: if we have an identity type $Id x y$ (propositionally equal) can we have a judgemetnal equality for the two terms in QIIT $x \equiv y$. \\
%Apparently, we don't setup our QIIT formulation in this way (I thought we need, but now I don't think we need it. Our QIIT formulation is purely in intensional MLTT now).\\ 
%But even if we do setup like that, I will not say a conflict is happening (because it is to our advantage if we have judgemental equal in metatheory). I will say it is a bit off as to how Coq is formulated (Coq is closer to an intensional MLTT, so no judgemental equal only propositional one). 
%}

% linkage, signature and compilation and seal
$$
\Rule[name=lsig/ept]
{""}
{\goodSig{\Gm}{\LSigEmp}{0}}
\quad
\Rule[name=ty/lkg]
{\goodSig{\Gm}{\lsig}{n}}
{\goodType{\Gm}{\TyLkg{\lsig}}{i}
\\ \goodType{\Gm}{\TyTkg{\lsig}}{}}
\quad
\Rule[name=Lnkg Compile]
{ \goodTerm{\Gamma}{o}{\cL \sigma} 
}
{
  \goodTerm{\Gamma}{\Tkg {o}}{\cC \sigma}
}
$$
$$
\Rule[name=Sig Add]
{\goodSig{\Gamma}{\sigma}{n} 
 \quad \goodType{\Gamma}{A}{}
 \quad \goodSeal{\Gamma}{f}{\sigma}{A}
 \quad \goodType{\Gamma, A}{T}{i}}
{\goodSig{\Gamma}{(\nu^+ \ \sigma \ \{f\} \ T)}{n+1}}
\quad
\Biggl( \boxed{\goodSeal{\Gamma}{s}{\sigma}{A}} 
:= \boxed{\goodTerm{\Gamma, \cC\sigma}{s}{A[p^1]}} \Biggr)
$$
$$
\Rule[name=Ept Lnkg]
{}
{\goodTerm{\Gamma}{\mu\cdot}{\cL \nu\cdot}}
\quad
\Rule[name=Lnkg Add]
{ \goodTerm{\Gamma}{o}{\cL \sigma} 
\quad \goodType{\Gamma}{A}{}
\quad  {\goodSeal{\Gamma}{f}{\sigma}{A}}
 \quad \boxed{\goodTerm{\Gamma, A}{t}{T}}
}
{\goodTerm{\Gamma}{(\mu^+ \ o \ \{f\} \ t)}{\cL(\nu^+ \ \sigma \ \{f\} \ T)}}
$$
$$
\Rule[name=lkg/proj]
{\goodTerm{\Gm}{o}{\TyLkg{\lsig}}}
{\goodTerm{\Gm}{\lkgproj{1} {o}}{\TyLkg {(\lsigproj{1} {\lsig})}}
\\ \boxed{\goodTerm{\Gm, \lsigprojT {\lsig}}{\lkgproj{2}{o}}{\lsigproj{2}{\lsig}}}
}
$$

We start with signature (in a brand new judgement $\goodSig{\Gamma}{\sigma}{n}$). A signature is a list of types contained by a \textit{linkage}\footnote{Following the convention by \citet{zm2017}, we name the meta-theoretical correspondents of \textit{family} as \textit{linkage}.}, tracking the length as superscript.   We will use $\{a : T_1, b : T_2, ..\}$ as a shorthand for signature with type $T_1, T_2, ..$ for readability.

\ruleref{lkg/proj} indicates how linkage captures the idea of \textit{code reuse}, as $\lkgproj{2}{-}$ will lead to the boxed term that is parametric on the surrounding linkage $\lsigprojT{\lsig}$ . To code reuse, we simply project out a field from the parent and add back to the children family.
\newpage 
Adding new types into signature \ruleref{Sig Add} requires a \textit{sealing judgement} $\goodSeal{\Gamma}{f}{\sigma}{A}$, which is an alias of a term judgement. Sealing is corresponding \textit{to indicating whether each field is overridable(extensible) or non-overridable (thus definitionally exposed) in the Coq Plugin}. 
Concretely, we encode our STLC example to show the motivation of the sealing.
%\YZ{Is sealing used only for late-binding the signatures of inductive types? Do I understand correctly that late binding of other kinds of fields (like subst) is achieved by the boxed premise in rule "LNKG ADD"?}\EDJreply{All late-bound are achieved by ``selective-abstraction'' imposed by sealing. For example, recursion will be consequently abstracted because of the abstraction of the inductive type it recurse on. 
%}\EDJreply{Sorry, I see what you mean. Yes. the box part is important for late-binding. But the functionality of sealing is always there. They are both required to achieve overriding. A illustrative example is, if the type is $S(\bot)$, then even with the boxed premise, it is still not overridable because the type itself restricts all the possibilities.}\YZreply{
%  By "selective abstraction", do you mean in the metatheory the programmer chooses what and how to abstract?
%}\EDJreply{Yes.}

In \cref{fig:sealing+linkage}, the "hiding" comment illustrates the late-bound prior fields for each fields.   For "tm", "ty" is late-bound and thus extending "ty" won't cause disruption on the inheritance of "tm". For "subst" itself, the definition of "tm"  cannot hidden for the sake of exhaustiveness checking. However, for each case handler inside "subst", "tm" is late-bound and thus each handler is inheritable.

Each hiding comment corresponds to the textual-aligned sealing derivation on RHS. For eligibility, curly brackets indiate a signature (omitting sealing at type level for readability) and square bracket indicate a nested sigma type, corresponding to sealing judgement $\goodSeal{}{s}{\sigma}{A}$.

For "tm", "ty" is hidden and thus the sealing "s₂" will make "ty" into an arbitrary type "𝕌" and expose its constructor "bool", "arrow", and etc. Then when constructing "o₂", the newly added field "tm" as "⋆ : 𝕎τ₂" is constructed in the context where "ty : 𝕌", ignoring its original eliminator. (Because eliminator is only possible when "ty : 𝕎τ₁"). On the other hand, for "subst", only "ty" is hidden and thus the sealing "s₃" leave "tm : 𝕎τ₂" untouched and "Wrec" becomes possible when constructing "o₃".

Basically, the hidden prior fields---inductive type are sealed into arbitrary type with ``constructors only'', where the recursion is constructed without sealing the inductive type.  Generally speaking, each extensible inductive type are hidden for all the other fields except for the recursion on itself. Furthermore, normal fields (e.g. "t1, t2") can expose its definition to all other fields in meta-theory by using singleton type. (e.g.  using "t1 : S(tvar 0)" instead of the current "t1 : tm"); and then overridable fields are hidden for normal fields (by sealing "t1 : S(tvar 0)" into "t1 : tm"). Generally speaking, anything with explicit "hiding" decoration can be understood as sealing.

Due to the definition of sealing, we can see the importance of compilation type, as the result of a sealing is usually a (nested) sigma type instead of a linkage type. For example, $\TyLkg{\{a : S(\bot), b : a \to \bot\}}$ can only be sealed into $\TyTkg{\{a : \cU, b : a \to \bot\}}$ but never $\TyLkg{\{a : \cU, b : a \to \bot\}}$. Because the latter cannot be inhabitted---by projecting the second field of the $ t : \TyLkg{\{a : \cU, b : a \to \bot\}}$, we can get a proof of bottom type. For simplicity, we make sealing a term in the context of compilation type instead of linkage type.

@@

% Inhertance judgement
Currently we can achieve inheritance by projecting out the field, and then add back the projection into the children family. That means the formulation of linkage already captures the concept of \textit{code reuse}.

However, in the plugin implementation, we use "Family"s to organize the code reuse. They are second class objects like Coq's "Module". What's more, a "Family" (with or without "extends" clause) can be considered as a standalone piece of code trying to \textit{inherit} an existent (but maybe empty) family. Thus we model them using \textit{inheritance judgement} $\goodInh{\Gamma}{h}{\sigma_1}{\sigma_2}$ instead of considering them as first class terms. The only way to use $h$ is by \ruleref{Inh} and transforming a linkage of signature $\sigma_1$ to another linkage of $\sigma_2$. Inheritance judgement can be inductively constructed via inheriting operations, overriding operations, extending operations and etc, corresponding to "Inherit", "Override" and "Extend" command  in our plugin implementation.
